Machine learning and essentialism

### 2023-02-01 ART_Sekrst/Sekrst.tex
Sekrst-Skansi article added

\begin{document}
\title{Machine learning and essentialism}
\maketitle

Abstract

Machine learning and essentialism have been connected in the past by various researchers, in order to state that the main paradigm in machine learning processes is equivalent to choosing the ``essential'' attributes for the machine to search for. Our goal in this paper is to show that there are connections between machine learning and essentialism, but only for some kinds of machine learning, and often not including deep learning methods. Similarity-based approaches, more connected to the overall prototype theory, spanning from psychology and linguistics, seem more suited for pattern recognition and complex deep-learning issues, while for classification problems, mostly for unsupervised learning, essentialism seems like the best choice. In order to illustrate the difference better, we will connect both paths to their sources in other disciplines and see how human psychology influences our decision in machine-learning modeling as well. This leads to a~philosophically very interesting consequence: even in the setting of supervised machine learning, essences are not present in data, but in targets, which in turn means that the categories which purport to be essences are in fact human-made, and hand-coded in the targets. The success of machine learning, therefore, does not give any substantial evidence for the independent existence of essential properties. Our stance here is to state that neither the existence nor the lack of ``essential'' properties in machine learning can lead to metaphysical, i.e., ontological claims.

Keywords

essentialism, machine learning, accidental properties, similarity-based approach, pattern recognition, modal necessity;

\section{Essential and accidental properties: introduction}
The purpose of this paper is to show that the existence of essential-like features in machine learning, or the lack of them, cannot provide an ontological commitment.\footnote{ The authors would like to thank the anonymous reviewers for their detailed analyses and insights.} Researchers have connected machine-learning practices with essentialist and anti-essentialist stances, but we feel that such claims ignore that both ``essentialist'' and ``anti-essentialist'' paradigms in machine learning are both influenced by human psychology and have no real consequence on the verification of whether there \textit{are} essential properties in nature or not.\footnote{ It is necessary to distinguish between ontological commitments regarding nature and ontological commitments that are necessary in every AI system (we call this difference an ontological gap). The former are the subject of this article, and the latter were analyzed by Krzanowski \& Polak
%\label{ref:RNDU3VddFbPO5}(2022a; 2022b).
\parencites**[][]{krzanowski_ontology_2022}[][]{krzanowski_meta-ontology_2022}.%
}

@@

\section{Philosophical essentialism}
An \textit{essential} property of an object is a~property that an object must have, while an \textit{accidental} one is the one the object happens could have, but that it could lack. That is, in modal terms,\footnote{ Standard modal characterizations have been disseminated with the works of Ruth Barcan Marcus and Saul Kripke. Kripke's work on semantics is taking the truth of a~formula relative to a~possible world, since its truth value depends on what is true in accessible world. See Barcan Marcus
%\label{ref:RNDktVIAPuvSw}(1993)
\parencite*[][]{marcus_modalities_1993} %
 for a~modality synthesis and Kripke 
%\label{ref:RND50iCIBiyUn}(1972)
\parencite*[][]{kripke_naming_1972} %
 for Kripke semantics.} we are talking about \textit{necessity} and \textit{possibility},\footnote{ There are, of course, differences between logical and (meta)physical possibilities. Something might not be a~logical contradiction, but still be (meta)physically impossible, i.e., not conforming to the laws of nature, for example, a~man travelling faster than the speed of light. The exact details of such classifications, especially between physical and metaphysical possibilities, are a~matter of debate.} respectively 
%\label{ref:RNDcC0OGIZEmV}(Robertson Ishii and Atkins, 2020):
\parencite[][]{robertson_ishii_essential_2020}: %
 an object is going to possess the essential property in all possible worlds, but for an accidental one, there is a~possible world in which an object lacks such a~property. \textit{Essentialism} is a~standpoint in which (at least) some objects have (at least some) essential properties 
%\label{ref:RNDS25JSRGYR9}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For example, an essential property of Socrates is to have originated from his parents but is not essential for him to have brown hair. An essential property of a~dog is certainly not brown hair since there are dogs of other colors. In philosophy, some essential properties are not a~matter of much debate. For example, a~dog had to have some biological origin. \textit{Canis canis} is also a~dog. But in order to pursue the matter further, there might be various objections to candidates for essential purposes.\footnote{ A~concise description of the debate is provided by Cartwright 
%\label{ref:RNDsX991rAklO}(1968, p.615):
\parencite*[][p.615]{cartwright_remarks_1968}: %
 ``What are the essential attributes of, say, Dancer's Image? No doubt it will be counted essential that he is a~horse and accidental that he was disqualified in this year's Kentucky Derby. But what of the attribute of being male, or of being a~thoroughbred, or of not being a~Clydesdale stallion? Here, I~suppose, essentialists may disagree. Indeed, a~reasonable essentialist might well take the position that these are hard cases that admit of no clear decision.}'' Often, a~dog is considered a~``wolf-like descendant'', where various breeds might not conform to this ideal, along with the notion of ``having an upturning tail''.

The former is the reason why there are various kinds of essentialism in philosophy. Standard Aristotelian essentialism also deals with necessities, and in his categories, he was researching properties that all the members of the category have in common, without which, they cannot be members of that category.\footnote{ For more details on Aristotelian essentialism, see Aristotle
%\label{ref:RNDTOuD48kM4B}(2014)
\parencite*[][]{aristotle_categories_2014} %
 and Matthews 
%\label{ref:RNDbP51aluDIe}(1990).
\parencite*[][]{matthews_aristotelian_1990}.%
} One of the most famous criticisms comes from Wittgenstein,\footnote{ See Cohen 
%\label{ref:RNDAJh4sLK2zw}(1968)
\parencite*[][]{cohen_wittgensteins_1968} %
 for more details.} who observed the debate from a~linguistic angle and stated that words can mean innumerable things depending on their use, paving the way to modern pragmatics. Probably the most common standpoint takes into account that both minimal and maximal essentialism apply. \textit{Maximal essentialism} states that all of any given object's properties are essential to it, while \textit{minimal essentialism} presupposes that there are no limits to the ways a~given object might have been different from its current actual state, and the only essential properties seem to be trivial ones, like ``being \textit{F}'' or ``being non-\textit{F}'' for any property \textit{F} and ``being self-identical'' 
%\label{ref:RNDBOk8l01X1n}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For the purpose of this paper, we will consider the most common stance as our starting point: maximal and minimal essentialism both hold. The mentioned doctrine that at least some objects have at least some essential properties is the most common one 
%\label{ref:RNDy9gdwlePak}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}.%
\footnote{ Explicitly stated by Mackie 
%\label{ref:RNDRNAllOD28Y}(2006).
\parencite*[][]{mackie_how_2006}. %
 For more details about various types of essentialism, see 
%\label{ref:RNDrEbONsf3qh}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}.%
}

Previous work on machine learning and essentialism dealt with various types of machine learning under the same hood, connecting them often to essentialist ideas. Works of Pelillo
%\label{ref:RNDoOwIXd5jk7}(2013),
\parencite*[][]{hutchison_how_2013}, %
 Pelillo and Scantamburlo 
%\label{ref:RNDXS1sLrFog8}(2013)
\parencite*[][]{hutchison_how_2013} %
 seem to be the most prominent ones. Tunç 
%\label{ref:RNDxxndDaar7v}(2015)
\parencite*[][]{tunc_semantics_2015} %
 follows Pelillo's 
%\label{ref:RNDyER8ceNEHr}(2013)
\parencite*[][]{hutchison_how_2013} %
 ideas but mostly focuses on epistemology and inductive inference, emphasizing abstracting, idealization, and theoretical variables in machine-learning research. Duin 
%\label{ref:RNDbLLJpi1ZqW}(2015)
\parencite*[][]{duin_dissimilarity_2015} %
 provides an anti-essentialist approach in pattern-recognition systems, claiming that in most of the applications in pattern recognition, there is no known, small set of essential features (a notion we agree with).\textcolor{black}{~Our goal is to show how various cases of essentialist-like and non-essentialist-like stances can be seen manifested in machine-learning choices, but that does not mean we are talking about real essentialist or anti-essentialist ontology.}

@@

In machine learning, data is usually split into \textit{training} and \textit{test} data, the same way a~student learns methods and approaches to some problems and gets previously unseen ones in an exam. Such an approach, compared to learning in the presence of a~supervisor or a~teacher, is called \textit{supervised learning}: an algorithm learns from \textit{labeled} data and is able to predict outcomes on previously unseen data. For example, if we had a~dataset consisting of various pictures of animals, and we wanted to train the algorithm to recognize cats, we would want it to be able to somehow point out what is \textit{essential} for an animal to be classified as a~cat. An important part of supervised learning is therefore the act of \textit{classification}: a~certain object of interest possesses or does not possess certain property, i.e., it is or is not a~member of a~class. A~certain image of a~dog might be marked as 98.6\% dog if it is very close to all of the properties that seem to be essential for classifying a~picture of an animal as a~dog. However, a~cat might have some properties, such as four legs and a~tail, but that would be a~low percentage. In another class of problems, there are \textit{regression} problems, in which the algorithm is predicting continuous values. For example, given previous real-estate prices in a~certain area, predict the prices for the next couple of years. Here, we would be dealing with real numbers instead of binary Boolean classifications.\footnote{ As a~side note, most of the algorithms do not predict using Boolean outcomes such as 0 or 1 for not being or being a~dog, but as a~matter of a~percentage. In such cases, we are effectively talking about fuzzy intervals.}

@@

Another type of machine learning, \textit{unsupervised learning}, handles various datasets without any explicit instructions or labels. That is, the learner receives unlabeled training data and makes predictions for all unseen points, and ``since, in general, no labeled example is available [...], it can be difficult to quantitatively evaluate the performance of a~learner''
%\label{ref:RNDCBD4vgLA3T}(Mohri, Rostamizadeh and Talwalkar, 2018, p.7).
\parencite[][p.7]{mohri_foundations_2018}. %
 Unsupervised learning encompasses a~broad definition of learning without labels or targets, but this broad definition begs the cognitive question of how we learn without feedback 
%\label{ref:RNDUR3el9pJJt}(Skansi, 2018, p.70).
\parencite[][p.70]{skansi_introduction_2018}. %
 In the previous case, in order for the computer to learn what is a~dog, we had to correctly label dogs or provide a~list of properties in other supervised examples. Here, a~computer is seemingly on its own: for example, neural networks\footnote{ See Skansi 
%\label{ref:RNDys19TmMXkr}(2018)
\parencite*[][]{skansi_introduction_2018} %
 for an introduction to deep learning.} tend to automatically find structures in the data by analyzing useful features. Data is often grouped into \textit{clusters}, and then it is easy to see the outliers, anomalies (for example, for fraud detection), associations (for recommender systems), and similar connections.

We might start to notice something interesting here. First, if we are telling the computer while we are labeling the data that something is or is not a~certain kind of object, we are effectively taking a~certain essentialist stance. Intuitively, there seems to be something \textit{essential} in all of the properties that make a~cat a~\textit{cat}. In various cases of supervised learning, we might list a~number of features that we could consider important. For example, my algorithm might be tracking pointy ears, four legs, two eyes, and fur. But such an algorithm might recognize dogs and rabbits as well but miss some dogs without pointy ears. And we are not even starting to talk about three-legged dogs and similar ``obviously'' accidental properties. Second, it all boils down to starting human decisions. This seems like a~trivial claim, and from a~description of supervised learning, it is rather intuitive. Blaming it on the data might seem like a~common excuse in machine learning, but recently, AI ethics has dwelled on the questions of initial data handling and responsibilities.\footnote{ A~famous example of an accidental algorithmic breach of ethics includes machine-learning racism in tagging black people as gorillas. See Zhang
%\label{ref:RNDS7hunknj5m}(2015).
\parencite*[][]{zhang_google_2015}.%
} However, why did we choose some features on top of others? The answer might lie in human psychology.

@@

Semic analysis was the first approach that aimed to find minimal units of meaning, which later developed into c\textit{omponential analysis} within the standard structuralist framework. In particular, Pottier
%\label{ref:RNDfcXAmU3MlY}(1964)
\parencite*[][]{pottier_vers_1964} %
 analyzed various types of chairs in order to find out what are the minimal features needed in order to distinguish between them. For example, they might have a~back side or not, might have arms or not, can be fixed or folding, can have one seat or several seats, etc. One can see that we are already dealing with both discrete and continuous values here. A~classic example in the componential analysis is how to describe various words for human beings in various stages of their lives, taking into account their gender. A~\textit{man} can be described as [-woman] and [+adult] or [+man] and [+adult]. Here we are dealing with Boolean man/female and adult/not adult, which does not reflect the fuzzy values of such categories, but structuralist linguistics was extremely focused on binary oppositions. Next, a~\textit{woman} would be [+woman] and [+adult] or [-man] and [+adult], a~\textit{girl} would be [+woman] but [-adult] or [-man] and [-adult], while a~\textit{boy} could be described as [-woman] and [-adult] or [+man] and [-adult].\footnote{ Such a~method was formed on the basis of Prague structuralist school dealing with phonology. A~phoneme has a~set of discrete properties, for example \textit{b} would be [+voiced], while \textit{p} would be [-voiced], but both would be [+labial] plosives.} Such an approach has been developed and changed but is still used in semantics, which, as one of its tasks, analyzes the internal structure of a~word by finding distinct and minimal components of meaning 
%\label{ref:RNDzgkSWwYznJ}(Palmer, 1981, p.108).
\parencite[][p.108]{palmer_semantics_1981}. %
 In such a~framework, we might differentiate our \textit{dog} from a~\textit{wolf} by finding distinct features. Both are certainly [+animal] and [+canine], but we might add [+domesticated] to the dog and [-domesticated] to the wolf. Such choices often seem arbitrary and there is no consensus on what the most basic properties of objects or classes of objects are, and it would also seem necessary to connect not only machine-learning feature selection with philosophy but linguistics and psychology as well.

@@

In his overview of pattern recognition, which is mostly unsupervised, Pelillo
%\label{ref:RND7VxQAwhbfg}(2013)
\parencite*[][]{hutchison_how_2013} %
 states that features are \textit{essential} properties. He reports Watanabe 
%\label{ref:RNDeIdvs8mrR3}(1985)
\parencite*[][]{watanabe_pattern_1985} %
 stating that ``under all works of pattern recognition lies tacitly the Aristotelian view that the world consists of a~discrete number of self-identical objects provided with, other than fleeting accidental properties, a~number of fixed or very slowly changing attributes. Some of these attributes, which may be called ‘features', determine the class to which the object belongs. Pellilo 
%\label{ref:RNDqHvs9V1I1J}(2013, p.2)
\parencite*[][p.2]{hutchison_how_2013} %
 reaffirms that the goal of a~pattern recognition algorithm is to discern ``the essences of a~category'' and that we should talk about an essentialist paradigm in machine learning. We have already mentioned Rosch's 
%\label{ref:RNDY1y7hcxV1q}(1973)
\parencite*[][]{rosch_natural_1973} %
 work on prototypes, which Pelillo 
%\label{ref:RNDODLfdxwA1F}(2013, p.2)
\parencite*[][p.2]{hutchison_how_2013} %
 uses to illustrate the ``multifaceted nature of real-world categories'' and emphasizes that for anti-essentialist stances, relations are in focus. That, of course, does not have to be the case, the main idea for anti-essentialism is to claim accidentality: there are possible worlds in which the object has the property in question and possible worlds in which it does not. But he does emphasize that the feature-based aspect is a~\textit{reductionist} position since objects are seen in isolation and overlook relational or contextual information 
%\label{ref:RNDCuNfcNeksb}(Pelillo, 2013, p.1).
\parencite[][p.1]{hutchison_how_2013}.%

The notion of a \textit{feature vector} is often used in machine learning: an \textit{n}}{-dimensional vector that serves a~purpose of a~collection of features. For example, just as a~red/green/blue combination will form a~single color, a~certain combination of features will be used in machine-learning tasks to better identify objects or predict values. Pelillo
%\label{ref:RNDpGLwZy2UZY}(2013, p.3)
\parencite*[][p.3]{hutchison_how_2013} %
 emphasizes that the community has focused on feature-vector representations, rather than on single, standalone features. In computer vision and pattern recognition, each object is described in terms of a~vector of numerical attributes and mapped to a~point in a~Euclidean vector space, so that the distances between the points reflect the similarities and dissimilarities between the respective objects 
%\label{ref:RNDN1XEVlzSff}(Pelillo, 2013, p.3).
\parencite[][p.3]{hutchison_how_2013}. %
 Pelillo emphasizes the recent trend in \textit{similarity-based techniques}, which are still not challenging the traditional paradigm but work with graphs or structural representations to find objects or values that seem to be closer according to some criterion 
%\label{ref:RNDIYspwzQPwH}(Pelillo, 2013, p.4).
\parencite[][p.4]{hutchison_how_2013}. %
 We have to note that such an approach is analogous to a~prototypical relationship, where members are grouped around a~prototype in a~certain graph-like manner. A~green apple is more similar to the prototype of a~red apple than a~red strawberry, and if such connections would be shown as a~weighted graph, then we would expect a~less expensive traversal to a~red apple.

>>>>

\begin{artengenv2auth}{Kristina Šekrst, Sandro Skansi}
	{Machine learning and essentialism}
	{Machine learning and essentialism}
	{Machine learning and essentialism}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{Machine learning and essentialism have been connected in the past by various researchers, in order to state that the main paradigm in machine learning processes is equivalent to choosing the ``essential'' attributes for the machine to search for. Our goal in this paper is to show that there are connections between machine learning and essentialism, but only for some kinds of machine learning, and often not including deep learning methods. Similarity-based approaches, more connected to the overall prototype theory, spanning from psychology and linguistics, seem more suited for pattern recognition and complex deep-learning issues, while for classification problems, mostly for unsupervised learning, essentialism seems like the best choice. In order to illustrate the difference better, we will connect both paths to their sources in other disciplines and see how human psychology influences our decision in machine-learning modeling as well. This leads to a~philosophically very interesting consequence: even in the setting of supervised machine learning, essences are not present in data, but in targets, which in turn means that the categories which purport to be essences are in fact human-made, and hand-coded in the targets. The success of machine learning, therefore, does not give any substantial evidence for the independent existence of essential properties. Our stance here is to state that neither the existence nor the lack of ``essential'' properties in machine learning can lead to metaphysical, i.e., ontological claims.}
	{essentialism, machine learning, accidental properties, similarity-based approach, pattern recognition, modal necessity.}
	{%
		{\flushright\subbold{Kristina Šekrst}\\\subsubsectit\small{University of Zagreb}\par}%
		{\flushright\subbold{Sandro Skansi}\\\subsubsectit\small{University of Zagreb}\par}%
	}

%\begin{document}
%\title{Machine learning and essentialism}
%\maketitle
%
%Abstract
%
%Machine learning and essentialism have been connected in the past by various researchers, in order to state that the main paradigm in machine learning processes is equivalent to choosing the ``essential'' attributes for the machine to search for. Our goal in this paper is to show that there are connections between machine learning and essentialism, but only for some kinds of machine learning, and often not including deep learning methods. Similarity-based approaches, more connected to the overall prototype theory, spanning from psychology and linguistics, seem more suited for pattern recognition and complex deep-learning issues, while for classification problems, mostly for unsupervised learning, essentialism seems like the best choice. In order to illustrate the difference better, we will connect both paths to their sources in other disciplines and see how human psychology influences our decision in machine-learning modeling as well. This leads to a~philosophically very interesting consequence: even in the setting of supervised machine learning, essences are not present in data, but in targets, which in turn means that the categories which purport to be essences are in fact human-made, and hand-coded in the targets. The success of machine learning, therefore, does not give any substantial evidence for the independent existence of essential properties. Our stance here is to state that neither the existence nor the lack of ``essential'' properties in machine learning can lead to metaphysical, i.e., ontological claims.
%
%Keywords
%
%essentialism, machine learning, accidental properties, similarity-based approach, pattern recognition, modal necessity.

\section{Essential and accidental properties: introduction}
The purpose of this paper is to show that the existence of essential-like features in machine learning, or the lack of them, cannot provide an ontological commitment.\footnote{The authors would like to thank the anonymous reviewers for their detailed analyses and insights.} Researchers have connected machine-learning practices with essentialist and anti-essentialist stances, but we feel that such claims ignore that both ``essentialist'' and ``anti-essentialist'' paradigms in machine learning are both influenced by human psychology and have no real consequence on the verification of whether there \textit{are} essential properties in nature or not.\footnote{It is necessary to distinguish between ontological commitments regarding nature and ontological commitments that are necessary in every AI system (we call this difference an ontological gap). The former are the subject of this article, and the latter were analyzed by Krzanowski \& Polak
%\label{ref:RNDU3VddFbPO5}(2022a; 2022b).
\parencites*[][]{krzanowski_ontology_2022}[][]{krzanowski_meta-ontology_2022}.%
}

@@

\section{Philosophical essentialism}
An \textit{essential} property of an object is a~property that an object must have, while an \textit{accidental} one is the one the object happens could have, but that it could lack. That is, in modal terms,\footnote{Standard modal characterizations have been disseminated with the works of Ruth Barcan Marcus and Saul Kripke. Kripke's work on semantics is taking the truth of a~formula relative to a~possible world, since its truth value depends on what is true in accessible world. See Barcan Marcus
%\label{ref:RNDktVIAPuvSw}(1993)
\parencite*[][]{marcus_modalities_1993} %
 for a~modality synthesis and Kripke 
%\label{ref:RND50iCIBiyUn}(1972)
\parencite*[][]{kripke_naming_1972} %
 for Kripke semantics.} we are talking about \textit{necessity} and \textit{possibility},\footnote{There are, of course, differences between logical and (meta)physical possibilities. Something might not be a~logical contradiction, but still be (meta)physically impossible, i.e., not conforming to the laws of nature, for example, a~man travelling faster than the speed of light. The exact details of such classifications, especially between physical and metaphysical possibilities, are a~matter of debate.} respectively 
%\label{ref:RNDcC0OGIZEmV}(Robertson Ishii and Atkins, 2020):
\parencite[][]{robertson_ishii_essential_2020}: %
 an object is going to possess the essential property in all possible worlds, but for an accidental one, there is a~possible world in which an object lacks such a~property. \textit{Essentialism} is a~standpoint in which (at least) some objects have (at least some) essential properties 
%\label{ref:RNDS25JSRGYR9}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For example, an essential property of Socrates is to have originated from his parents but is not essential for him to have brown hair. An essential property of a~dog is certainly not brown hair since there are dogs of other colors. In philosophy, some essential properties are not a~matter of much debate. For example, a~dog had to have some biological origin. \textit{Canis canis} is also a~dog. But in order to pursue the matter further, there might be various objections to candidates for essential purposes.\footnote{A~concise description of the debate is provided by Cartwright 
%\label{ref:RNDsX991rAklO}(1968, p.615):
\parencite*[][p.615]{cartwright_remarks_1968}: %
 ``What are the essential attributes of, say, Dancer's Image? No doubt it will be counted essential that he is a~horse and accidental that he was disqualified in this year's Kentucky Derby. But what of the attribute of being male, or of being a~thoroughbred, or of not being a~Clydesdale stallion? Here, I~suppose, essentialists may disagree. Indeed, a~reasonable essentialist might well take the position that these are hard cases that admit of no clear decision.}'' Often, a~dog is considered a~``wolf-like descendant'', where various breeds might not conform to this ideal, along with the notion of ``having an upturning tail''.

The former is the reason why there are various kinds of essentialism in philosophy. Standard Aristotelian essentialism also deals with necessities, and in his categories, he was researching properties that all the members of the category have in common, without which, they cannot be members of that category.\footnote{For more details on Aristotelian essentialism, see Aristotle
%\label{ref:RNDTOuD48kM4B}(2014)
\parencite*[][]{aristotle_categories_2014} %
 and Matthews 
%\label{ref:RNDbP51aluDIe}(1990).
\parencite*[][]{matthews_aristotelian_1990}.%
} One of the most famous criticisms comes from Wittgenstein,\footnote{See Cohen 
%\label{ref:RNDAJh4sLK2zw}(1968)
\parencite*[][]{cohen_wittgensteins_1968} %
 for more details.} who observed the debate from a~linguistic angle and stated that words can mean innumerable things depending on their use, paving the way to modern pragmatics. Probably the most common standpoint takes into account that both minimal and maximal essentialism apply. \textit{Maximal essentialism} states that all of any given object's properties are essential to it, while \textit{minimal essentialism} presupposes that there are no limits to the ways a~given object might have been different from its current actual state, and the only essential properties seem to be trivial ones, like ``being \textit{F}'' or ``being non-\textit{F}'' for any property \textit{F} and ``being self-identical'' 
%\label{ref:RNDBOk8l01X1n}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For the purpose of this paper, we will consider the most common stance as our starting point: maximal and minimal essentialism both hold. The mentioned doctrine that at least some objects have at least some essential properties is the most common one 
%\label{ref:RNDy9gdwlePak}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}.%
\footnote{Explicitly stated by Mackie 
%\label{ref:RNDRNAllOD28Y}(2006).
\parencite*[][]{mackie_how_2006}. %
 For more details about various types of essentialism, see 
%\label{ref:RNDrEbONsf3qh}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}.%
}

Previous work on machine learning and essentialism dealt with various types of machine learning under the same hood, connecting them often to essentialist ideas. Works of Pelillo
%\label{ref:RNDoOwIXd5jk7}(2013),
\parencite*[][]{pelillo_introduction_2013}, %
 Pelillo and Scantamburlo 
%\label{ref:RNDXS1sLrFog8}(2013)
\parencite*[][]{hutchison_how_2013} %
 seem to be the most prominent ones. Tunç 
%\label{ref:RNDxxndDaar7v}(2015)
\parencite*[][]{tunc_semantics_2015} %
 follows Pelillo's 
%\label{ref:RNDyER8ceNEHr}(2013)
\parencite*[][]{pelillo_introduction_2013} %
 ideas but mostly focuses on epistemology and inductive inference, emphasizing abstracting, idealization, and theoretical variables in machine-learning research. Duin 
%\label{ref:RNDbLLJpi1ZqW}(2015)
\parencite*[][]{duin_dissimilarity_2015} %
 provides an anti-essentialist approach in pattern-recognition systems, claiming that in most of the applications in pattern recognition, there is no known, small set of essential features (a notion we agree with). Our goal is to show how various cases of essentialist-like and non-essentialist-like stances can be seen manifested in machine-learning choices, but that does not mean we are talking about real essentialist or anti-essentialist ontology.

@@

In machine learning, data is usually split into \textit{training} and \textit{test} data, the same way a~student learns methods and approaches to some problems and gets previously unseen ones in an exam. Such an approach, compared to learning in the presence of a~supervisor or a~teacher, is called \textit{supervised learning}: an algorithm learns from \textit{labeled} data and is able to predict outcomes on previously unseen data. For example, if we had a~dataset consisting of various pictures of animals, and we wanted to train the algorithm to recognize cats, we would want it to be able to somehow point out what is \textit{essential} for an animal to be classified as a~cat. An important part of supervised learning is therefore the act of \textit{classification}: a~certain object of interest possesses or does not possess certain property, i.e., it is or is not a~member of a~class. A~certain image of a~dog might be marked as 98.6\% dog if it is very close to all of the properties that seem to be essential for classifying a~picture of an animal as a~dog. However, a~cat might have some properties, such as four legs and a~tail, but that would be a~low percentage. In another class of problems, there are \textit{regression} problems, in which the algorithm is predicting continuous values. For example, given previous real-estate prices in a~certain area, predict the prices for the next couple of years. Here, we would be dealing with real numbers instead of binary Boolean classifications.\footnote{As a~side note, most of the algorithms do not predict using Boolean outcomes such as 0 or 1 for not being or being a~dog, but as a~matter of a~percentage. In such cases, we are effectively talking about fuzzy intervals.}

@@

Another type of machine learning, \textit{unsupervised learning}, handles various datasets without any explicit instructions or labels. That is, the learner receives unlabeled training data and makes predictions for all unseen points, and ``since, in general, no labeled example is available [...], it can be difficult to quantitatively evaluate the performance of a~learner''
%\label{ref:RNDCBD4vgLA3T}(Mohri, Rostamizadeh and Talwalkar, 2018, p.7).
\parencite[][p.7]{mohri_foundations_2018}. %
 Unsupervised learning encompasses a~broad definition of learning without labels or targets, but this broad definition begs the cognitive question of how we learn without feedback 
%\label{ref:RNDUR3el9pJJt}(Skansi, 2018, p.70).
\parencite[][p.70]{skansi_introduction_2018}. %
 In the previous case, in order for the computer to learn what is a~dog, we had to correctly label dogs or provide a~list of properties in other supervised examples. Here, a~computer is seemingly on its own: for example, neural networks\footnote{See Skansi 
%\label{ref:RNDys19TmMXkr}(2018)
\parencite*[][]{skansi_introduction_2018} %
 for an introduction to deep learning.} tend to automatically find structures in the data by analyzing useful features. Data is often grouped into \textit{clusters}, and then it is easy to see the outliers, anomalies (for example, for fraud detection), associations (for recommender systems), and similar connections.

We might start to notice something interesting here. First, if we are telling the computer while we are labeling the data that something is or is not a~certain kind of object, we are effectively taking a~certain essentialist stance. Intuitively, there seems to be something \textit{essential} in all of the properties that make a~cat a~\textit{cat}. In various cases of supervised learning, we might list a~number of features that we could consider important. For example, my algorithm might be tracking pointy ears, four legs, two eyes, and fur. But such an algorithm might recognize dogs and rabbits as well but miss some dogs without pointy ears. And we are not even starting to talk about three-legged dogs and similar ``obviously'' accidental properties. Second, it all boils down to starting human decisions. This seems like a~trivial claim, and from a~description of supervised learning, it is rather intuitive. Blaming it on the data might seem like a~common excuse in machine learning, but recently, AI ethics has dwelled on the questions of initial data handling and responsibilities.\footnote{A~famous example of an accidental algorithmic breach of ethics includes machine-learning racism in tagging black people as gorillas. See Zhang
%\label{ref:RNDS7hunknj5m}(2015).
\parencite*[][]{zhang_google_2015}.%
} However, why did we choose some features on top of others? The answer might lie in human psychology.

@@

Semic analysis was the first approach that aimed to find minimal units of meaning, which later developed into \textit{componential analysis} within the standard structuralist framework. In particular, Pottier
%\label{ref:RNDfcXAmU3MlY}(1964)
\parencite*[][]{pottier_vers_1964} %
 analyzed various types of chairs in order to find out what are the minimal features needed in order to distinguish between them. For example, they might have a~back side or not, might have arms or not, can be fixed or folding, can have one seat or several seats, etc. One can see that we are already dealing with both discrete and continuous values here. A~classic example in the componential analysis is how to describe various words for human beings in various stages of their lives, taking into account their gender. A~\textit{man} can be described as [-woman] and [+adult] or [+man] and [+adult]. Here we are dealing with Boolean man/female and adult/not adult, which does not reflect the fuzzy values of such categories, but structuralist linguistics was extremely focused on binary oppositions. Next, a~\textit{woman} would be [+woman] and [+adult] or [-man] and [+adult], a~\textit{girl} would be [+woman] but [-adult] or [-man] and [-adult], while a~\textit{boy} could be described as [-woman] and [-adult] or [+man] and [-adult].\footnote{Such a~method was formed on the basis of Prague structuralist school dealing with phonology. A~phoneme has a~set of discrete properties, for example \textit{b} would be [+voiced], while \textit{p} would be [-voiced], but both would be [+labial] plosives.} Such an approach has been developed and changed but is still used in semantics, which, as one of its tasks, analyzes the internal structure of a~word by finding distinct and minimal components of meaning 
%\label{ref:RNDzgkSWwYznJ}(Palmer, 1981, p.108).
\parencite[][p.108]{palmer_semantics_1981}. %
 In such a~framework, we might differentiate our \textit{dog} from a~\textit{wolf} by finding distinct features. Both are certainly [+animal] and [+canine], but we might add [+domesticated] to the dog and [-domesticated] to the wolf. Such choices often seem arbitrary and there is no consensus on what the most basic properties of objects or classes of objects are, and it would also seem necessary to connect not only machine-learning feature selection with philosophy but linguistics and psychology as well.

@@

In his overview of pattern recognition, which is mostly unsupervised, Pelillo
%\label{ref:RND7VxQAwhbfg}(2013)
\parencite*[][]{pelillo_introduction_2013} %
 states that features are \textit{essential} properties. He reports Watanabe 
%\label{ref:RNDeIdvs8mrR3}(1985)
\parencite*[][]{watanabe_pattern_1985} %
 stating that ``under all works of pattern recognition lies tacitly the Aristotelian view that the world consists of a~discrete number of self-identical objects provided with, other than fleeting accidental properties, a~number of fixed or very slowly changing attributes. Some of these attributes, which may be called ‘features', determine the class to which the object belongs. Pellilo 
%\label{ref:RNDqHvs9V1I1J}(2013, p.2)
\parencite*[][p.2]{pelillo_introduction_2013} %
 reaffirms that the goal of a~pattern recognition algorithm is to discern ``the essences of a~category'' and that we should talk about an essentialist paradigm in machine learning. We have already mentioned Rosch's 
%\label{ref:RNDY1y7hcxV1q}(1973)
\parencite*[][]{rosch_natural_1973} %
 work on prototypes, which Pelillo 
%\label{ref:RNDODLfdxwA1F}(2013, p.2)
\parencite*[][p.2]{pelillo_introduction_2013} %
 uses to illustrate the ``multifaceted nature of real-world categories'' and emphasizes that for anti-essentialist stances, relations are in focus. That, of course, does not have to be the case, the main idea for anti-essentialism is to claim accidentality: there are possible worlds in which the object has the property in question and possible worlds in which it does not. But he does emphasize that the feature-based aspect is a~\textit{reductionist} position since objects are seen in isolation and overlook relational or contextual information 
%\label{ref:RNDCuNfcNeksb}(Pelillo, 2013, p.1).
\parencite[][p.1]{pelillo_introduction_2013}.%

The notion of a \textit{feature vector} is often used in machine learning: an \textit{n}-dimensional vector that serves a~purpose of a~collection of features. For example, just as a~red/green/blue combination will form a~single color, a~certain combination of features will be used in machine-learning tasks to better identify objects or predict values. Pelillo
%\label{ref:RNDpGLwZy2UZY}(2013, p.3)
\parencite*[][p.3]{pelillo_introduction_2013} %
 emphasizes that the community has focused on feature-vector representations, rather than on single, standalone features. In computer vision and pattern recognition, each object is described in terms of a~vector of numerical attributes and mapped to a~point in a~Euclidean vector space, so that the distances between the points reflect the similarities and dissimilarities between the respective objects 
%\label{ref:RNDN1XEVlzSff}(Pelillo, 2013, p.3).
\parencite[][p.3]{pelillo_introduction_2013}. %
 Pelillo emphasizes the recent trend in \textit{similarity-based techniques}, which are still not challenging the traditional paradigm but work with graphs or structural representations to find objects or values that seem to be closer according to some criterion 
%\label{ref:RNDIYspwzQPwH}(Pelillo, 2013, p.4).
\parencite[][p.4]{pelillo_introduction_2013}. %
 We have to note that such an approach is analogous to a~prototypical relationship, where members are grouped around a~prototype in a~certain graph-like manner. A~green apple is more similar to the prototype of a~red apple than a~red strawberry, and if such connections would be shown as a~weighted graph, then we would expect a~less expensive traversal to a~red apple.
